\section{Preliminaries}

\subsection{Notation}

The following table specifies the symbols and notation used in this document.

\begin{center}
\begin{tabular}{c l}
    \hex{1337} & hexadecimal number
    \\ $\bot$ & verification failure
    \\ $x \cat y$ & concatenation of bitstrings $x$ and $y$
    \\ $x \xor y$ & bitwise addition of bitstrings $x$ and $y$
    \\ $x \ror{32} n$ & 32-bit word $x$ rotated right by $n$ bits
    \\ $x \ror{64} n$ & 64-bit word $x$ rotated right by $n$ bits
\end{tabular}
\end{center}

\subsection{Symmetric encryption}

In communication, it is often desirable to keep one's messages hidden from third
parties. This property is called \emph{confidentiality}. In more precise terms,
confidentiality means that a message is transformed in such a way that
authorized parties can recover the original message from it, while unauthorized
parties can not.

In symmetric encryption, the authorized parties are defined as
the parties that know some secret information, called the \emph{key}. The
original message is called the \emph{plaintext} and the transformed message is
called the \emph{ciphertext}. The transformation from plaintext to ciphertext is
called \emph{encryption}, while the transformation back is called
\emph{decryption}. When the same secret key is needed for both encryption and
decryption, it is called \emph{symmetric encryption}. Because the key is needed
for decryption, unauthorized parties are prevented from recovering the
plaintext.

\subsubsection{Nonces}

While unauthorized parties are unable to decrypt ciphertexts, similar plaintext
may result in similar ciphertexts, and unauthorized parties may therefore detect
when similar messages are sent. In order to prevent this, during every
encryption, a different number is used to modify the resulting ciphertext. It is
needed again during decryption in order to revert that modification. Because a
different number is used every time, the resulting ciphertexts will also differ.

This number is called the \emph{nonce}, short for number used once. If it is
used more than once, unauthorized parties may be able to infer something about
the difference or similarity of the messages it was used for. Usually, this
number is made from a sufficiently large amount of random bits that the chance
of it occuring twice is negligable. When it is chosen randomly, it must be
attached to the ciphertext unencrypted, as decryption will fail without it.

If there is no space to attach the nonce to the ciphertext, it is not possible
to use a random nonce. In this case, information that is available during both
the encryption and decryption should be used to construct a unique nonce. For
example, when adding encryption support to existing filesystems, there may be no
space for nonces in the existing data structures and these probably need to stay
compatible with earlier versions. In this case, a nonce for a block can be
constructed by combining the inode and offset within the file into a nonce. With
this construction, uniqueness cannot be guaranteed, so security is reduced, but
if encryption is otherwise not possible, it is still an improvement over
unencrypted data or encryption without nonces.

\subsubsection{Authentication}

Although unauthorized parties cannot recover the plaintext from the ciphertext,
they may be able to modify the ciphertext, resulting in a modified plaintext
after decryption. It is not always possible to prevent such modifications, but
it is possible to detect them. The property that it is detectable whether a
message has been changed, is called \emph{integrity}.

Unauthorized parties may also attempt to construct messages from scratch.
Because the key is needed for encryption, they will be unable to encrypt a
specific plaintext, however, they will be able to send specific ciphertexts,
even if they don't know what plaintexts corresponds to them. In order to prevent
this, it is desirable for the receiving party to be able to verify that a
message comes from an authorized party. Together with integrity, this property
is called \emph{authenticity}. Encryption that provides both confidentiality and
authenticity to the plaintext is called \emph{authenticated encryption} or
\emph{AE}.

Authenticated encryption is usually implemented by generating an
\emph{authentication tag}. Just like the ciphertext, this tag is based on the
key, the nonce and the plaintext. This tag is needed again during decryption.
The decryption algorithm checks if the tag is correct and only returns a
plaintext if it is. If the ciphertext is modified, it will decrypt to a
different plaintext from the one used to generate the tag and cause the
decryption to fail. Because the tag also depends on the key, valid tags cannot
be created without it.

\subsubsection{Associated data}

Even when confidentiality and authenticity are assured, unauthorized parties may
still repeat a message they have seen before in a different context. To prevent
this, some data about the context in which a message is allowed to appear can be
associated with it. This is one of the uses of \emph{associated data}, which is
defined as information that requires authentication but not confidentiality.
Just like nonces, this \emph{associated data} is used during encryption to
modify the resulting ciphertext and is needed during decryption in order to
revert that modification. It does not need to be stored or sent with the
ciphertext if it can be inferred during both encryption and decryption.

Associated data can prevent a ciphertext from being decrypted in unintended
contexts. An authenticated encryption scheme that supports this is called
\emph{authenticated encryption with associated data} or \emph{AEAD}.

\subsubsection{Formal definition}

Security of encryption systems is often defined with respect to a security
parameter. The \emph{security parameter} specifies how much effort is required
to break the security properties of the system. The security parameter is
expressed in bits and is equivelant to the 2-log of the number of operations
needed to break the security claim. A system is secure when no algorithm exists
that can break any of its security requirements in fewer operation than
specified by its security parameter.

An authenticated encryption with associated data scheme is defined as a tuple of
functions $(E, D)$ with the following properties:

$E$ is a function that takes a key, a nonce, some associated data and a message
and produces a ciphertext and an authentication tag.
\begin{equation}
    E \colon K \times N \times A \times M \to C \times T
\end{equation}

$D$ is a function that takes a key, a nonce, some associated data, a ciphertext
and an authentication tag and produces either a failure or a message.
\begin{equation}
    D \colon K \times N \times A \times C \times T \to M \cup \{ \bot \}
\end{equation}

Let $k$ be a key, $n$ be a nonce, $a$ be any associated data and $m$ be a
message. The result of decryption after encryption with the same nonce, key and
associated data is the original message. In particular, this means decryption
never returns $\bot$ when supplied with valid encryption results.
\begin{equation}
    D(k, n, a, E(k, n, a, m)) = m
\end{equation}

% TODO: No algorithm constructs valid decryption inputs different from ones seen
% before without the key.
% TODO: No algorithm decrypts without the correct key, nonce, tag, associated
% data and ciphertext.
% TODO: Different nonces imply unrelated ciphertexts
% TODO: Incorrect tag, key or associated data implies decryption failure

Let $k$ be a key, $n$ be a nonce, $a$ be any associated data and $m$ be a
message and let $c, t = E(k, n, a, m)$. Authenticity means the probability is
negligable that a decryption succeeds for a different ciphertext and the same
nonce, key, associated data and tag.

\begin{equation}
    P(D(k, n, a, c', t) \neq \bot \mid c \neq c') < \frac{1}{2^S}
\end{equation}

\subsection{Ascon internals}

Ascon~\cite{ascon} is an authenticated encryption cipher designed for use in
resource-constrained environments, like embedded devices. It has an internal
state of just 320 bits, which can be kept in registers on most architectures.
This ensures moving data between registers and memory is kept to a minimum,
which is important, as embedded devices usually do not have the same amount of
cache available as larger systems.

\subsubsection{Mode of operation}

Ascon aims to provide 128 bits of security. To that end, its key, nonce, and
authentication tag are 128 bits in size each. The plaintext, ciphertext and
associated data can all be of any length and are processed in \emph{blocks}.
There are multiple variants of Ascon, we implement two of them: Ascon-128, which
processes 64-bit blocks and Ascon-128a, which processes 128-bit blocks.

Ascon uses a 320-bit state and a permutation that mixes the state in a way that
is hard to reverse. This permutation consists of a transformation that is
applied in multiple rounds, each with a different round constant. Both variants
of Ascon use permutation $p^a$, which consists of 12 rounds, during
initialization and finalization. After processing each block, Ascon uses
permutation $p^b$, which consists of 6 rounds in Ascon-128 and 8 rounds in
Ascon-128a. Figure~\ref{encdec} gives an overview of encryption and decryption.

\input{encdec}

Encryption starts with the initialization of the state. The state is initialized
to the concatenation of an initialization vector, the key and the nonce. The
initialization vector encodes the parameters of the variant of Ascon that it is
used for. This initial state is then passed through the first permutation,
$p^b$. Finally, the key is bitwise added to the least significant bits of the
state again.

After initialization, the associated data is mixed into the state. If there is
no associated data, the state is left unmodified. Otherwise, the associated data
is padded and split into blocks. Each of the blocks is bitwise added to the most
significant bits of the state, followed by applying the second permutation,
$p^b$. At the end, the least significant bit is inverted. This is done to
separate the associated data from the plaintext, which comes next.

The plaintext is padded and split into blocks in the same way as the associated
data. Each block of plaintext is bitwise added to the most significant bits of
the state. The result of this bitwise addition also forms the corresponding
block of the ciphertext. After adding each block, except for the last,
permutation $p^b$ is applied again.

Finalization begins with bitwise adding the key to the most significant bits of
the state that were not used to add the plaintext. After that, the permutation
$p^a$ is applied again and finally, the key is bitwise added to the least
significant bits of the state again. The result of this last operation is used
as the authentication tag.

Decryption is identical to encryption, except the ciphertext is processed in a
slightly different way from the plaintext: Each block of the ciphertext is again
bitwise added to the most significant bits of the state to form the plaintext,
but instead of using this result in the state, the most significant bits of the
state are replaced with the ciphertext before the bitwise addition.

\subsubsection{Permutation}

Ascon's main component is a permutation consisting of three phases, which are
applied in several rounds. Ascon-128 uses 6 rounds to process blocks of 64 bits
at a time. Ascon-128a uses 8 rounds to process blocks of 128 bits at a time.
Both use keys of 128 bits and 12 rounds to initialize and finalize the state. A
round consists of the following three phases: The addition of the round
constant, the substitution layer and the linear diffusion layer. Each of these
phases modify the internal state in a different way. The internal state consists
of 320 bits, logically split into five 64-bit state words called $x_0$, $x_1$,
$x_2$, $x_3$ and $x_4$.

The round constant is a single byte that changes from round to round and is
added bitwise to the least significant 8 bits of $x_2$. It ensures rounds are
not all identical. The final round constant is always \hex{4b}. The round
constant changes linearly, decreasing by \hex{f} every round. This means
the round constant can be computed easily based on the number of rounds that are
left.

The substitution layer applies a 5-bit lookup function in parallel to the five
state words. It provides non-linear mixing between the five state words. It is
usually implemented as a sequence of bitwise operations which compute the
substitution on five machine words in parallel.

The linear diffusion layer provides linear mixing of the bits within each state
word. It consists of a bitwise addition of the original state word with the same
state word rotated by two different amounts. Each of the five state words uses
different rotation amounts.

\subsection{RISC-V}
RISC-V~\cite{riscv} is a hardware instruction set architecture which aims to be
completely open: Anyone is free to create custom variations and implementations.
It is also designed to be modular: A basic RISC-V processor starts with just the
base integer instruction set \texttt{I}, using either 32-bit or 64-bit machine
words. Several extensions are available which offer different common processor
features, such as support for floating point operations. Manufacturers can
choose to implement any combination of these extensions and will usually pick
extensions based on the intended application.

In this thesis, we will use the HiFive1 development board created by SiFive. It
features a \texttt{RV32IMAC} processor. The \texttt{RV32I} stands for the RISC-V
32-bit base integer instruction set. The letters that follow stand for the
extensions that are implemented in this processor: \texttt{M} stands for the
multiplication extensions which offers integer multiplication and division
instructions, \texttt{A} stands for the atomic memory access extension and
\texttt{C} stands for the compressed instruction set extension which can be used
to reduce code size.

For Ascon, we won't need multiplication instructions or atomic memory access.
The compressed instructions are used transparently by the assembler: Each
compressed instruction corresponds to a single full size instruction and
whenever a compressed instruction is available, it will be used instead of a
full size instruction. We would like to use the bit manipulation extension, but
it is not available on this board and may not be available on other embedded
RISC-V systems, so we optimize Ascon for RISC-V processors without it.

\emph{RISC} stands for Reduced Instruction Set Computer, as opposed to CISC,
which stands for Complex Instruction Set Computer. RISC architectures aim to
have a small set of simple and general instructions, while CISC architectures
aim to provide complex instructions which can do multiple things at once. An
example of this is that the CISC x86 architecture allows memory access as part
of almost every instruction, while RISC architectures usually have separate
memory access instructions.

RISC-V takes RISC to an extreme, it is much more reduced than for example ARM,
which is often called a RISC architecture as well. For example, RISC-V does not
save carries for arithmetic operations and it only has about 40 instruction in
the base integer instruction set, all of which do only one thing.
Figure~\ref{opcodes} shows the eight instructions needed for the inner loop of
Ascon.

\begin{figure}
\begin{tabular}{l p{0.77\linewidth}}
       \texttt{and r, a, b} & Store the bitwise and of registers~\texttt{a} and
        \texttt{b} in register~\texttt{r}.
    \\ \texttt{or r, a, b} & Store the bitwise or of registers~\texttt{a} and
        \texttt{b} in register~\texttt{r}.
    \\ \texttt{xor r, a, b} & Store the bitwise addition of registers~\texttt{a}
        and \texttt{b} in register~\texttt{r}. This operation is also known as
        bitwise exclusive or.
    \\ \texttt{not r, a} & Store the bitwise inversion of register~\texttt{a} in
        register~\texttt{r}.
    \\ \texttt{addi r, a, b} & Store the sum of the value of register~\texttt{a}
        and constant~\texttt{b} in register~\texttt{r}.
    \\ \texttt{srli r, a, b} & Store the value of register~\texttt{a} shifted
        right by constant amount~\texttt{b} in register~\texttt{r}.
    \\ \texttt{ssli r, a, b} & Store the value of register~\texttt{a} shifted
        left by constant amount~\texttt{b} in register~\texttt{r}.
    \\ \texttt{bne a, b, l} & Jump to the label~\texttt{l} if the value of
        register~\texttt{a} is not the same as the value of register~\texttt{b}.
\end{tabular}

\caption{The RISC-V instructions used in the inner loop of Ascon.}

\label{opcodes}
\end{figure}
